---
title: "R Notebook"
output: html_notebook
---
Factor analysis of clinical questionnaire, PTSD data set

```{r}
rm(list = ls())
```


**LOAD PACKAGE** 
```{r}
# psych package
# install.packages('psych')
library(psych)

# MVN package for multivariate normality test
# install.packages("MVN")
library(MVN)

# nFactors package 
# install.packages('nFactors')
library(nFactors)

#install.packages('GPArotation')
library(GPArotation)

library(ggplot2)

# install.packages("PerformanceAnalytics")
library(PerformanceAnalytics)

# install.packages('xlsx')
# library(xlsx)
```

**LOAD DATA**
```{r}
tb = read.csv("D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/question scores EFA_09152018.csv", header = TRUE)

# exclude those who have incomplete data 
exclude = c(117, 1285, 100, 102, 50, 3) 

# find out who are remitted PTSD
remitted = tb$id[tb$group=='R']

```

**SELECT SUBJECT**  
Choice 1. include only control and PTSD, but also include women (should be 59 participants)
```{r}
tb2faGain <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]

tb2faLoss <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]
```

Choice 2. include only control and PTSD, (should be 55 participants)
```{r}
tb2faGain <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                & tb$isMale==1 & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]

tb2faLoss <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                & tb$isMale==1 & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]
```

Choice 3. include control, PTSD, remitted PTSD (should be 69 participants)
```{r}
tb2faGain <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
            & tb$isMale==1 & !is.element(tb$id,exclude),]

tb2faLoss <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                & tb$isMale==1 & !is.element(tb$id,exclude),]
```

Display column names of data table, make sure to check if the right columns are selected to do PCA
```{r}
names(tb2faGain)
```

**Prepare data for PCA**  
Organize into a seperate array
```{r}
array2fa <- tb2faGain[,c(34:39, 46, 47:49, 54)] # 11 questionnaitre
colnames(array2fa) = c("BDI", "CAPS-ReExp","CAPS-Avoid","CAPS-Numb","CAPS-DysA","CAPS-AnxA",
                       "CTQ","STAI-1","STAI-2","CES","DES")
```

Test multivariate normality
```{r}
mvn(array2fa)
```

Visualize data structure
```{r}
cor.plot(array2fa, numbers = TRUE, main="11 Quest Scores", xlas=2)
chart.Correlation(array2fa,
                  histogram = TRUE,
                  method = 'pearson')

# pairs.panels(array2fa,pch='.')
```

Z-Score
```{r}
# generate z-scores for variable A using the scale() function
zscored <- scale(array2fa, center = TRUE, scale = TRUE)

```

**Do PCA**  
There are two functions to use, princomp and prcomp
```{r}
# pcaresult <- princomp(zscored)
pcaresult <- prcomp(array2fa, retx = TRUE, center = TRUE, scale. = TRUE)
summary(pcaresult)
```
Plot variance explained
```{r}
plot(pcaresult)
```

Plot loadings
```{r}
# plot loadings
# pcaload <- pcaresult$loadings[,]
pcaload <- pcaresult$rotation[,]
barplot(pcaload[,1], main="Component 1", ylab="Loadings", las=2)
barplot(pcaload[,2], main="Component 2", ylab="Loadings", las=2)
barplot(pcaload[,3], main="Component 3", ylab="Loadings", las=2)

```

Get rotated values (component scores)
```{r}
# rotated values
score = pcaresult$x
```

**Save PCA output**  
Combine component scores into data
```{r}
# combine component into data
tb2faGain$comp1 = score[,1]
tb2faGain$comp2 = score[,2]
tb2faGain$comp3 = score[,3]

tb2faLoss$comp1 = score[,1]
tb2faLoss$comp2 = score[,2]
tb2faLoss$comp3 = score[,3]
```

Save data
```{r}
# save data
write.csv(tb2faGain,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score gain_femaleIn_noRemitted_01172019.csv")
write.csv(tb2faLoss,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score loss_femaleIn_noRemitted_01172019.csv")
```

**Caculate the components for remitted PTSD subjects**  
If excluding females
```{r}
# select remitted subjects only
tb2faGainRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                        & tb$isMale==1 & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]

tb2faLossRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                        & tb$isMale==1 & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]
```

If including females
```{r}
# select remitted subjects only
tb2faGainRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                        & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]

tb2faLossRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                        & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]

```

Calculate the components
```{r}
scoreRemitted = scale(tb2faGainRemitted[,c(34:39, 46, 47:49, 54)], center= pcaresult$center) %*% pcaload
```

Combine component score into data
```{r}
tb2faGainRemitted$comp1 = scoreRemitted[,1]
tb2faGainRemitted$comp2 = scoreRemitted[,2]
tb2faGainRemitted$comp3 = scoreRemitted[,3]

tb2faLossRemitted$comp1 = scoreRemitted[,1]
tb2faLossRemitted$comp2 = scoreRemitted[,2]
tb2faLossRemitted$comp3 = scoreRemitted[,3]

```

Save data
```{r}

write.csv(tb2faGainRemitted,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score gain_femaleIn_Remitted_01172019.csv")
write.csv(tb2faLossRemitted,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score loss_femaleIn_Remitted_01172019.csv")

```





When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
