---
title: "R Notebook"
output: html_notebook
---
Factor analysis of clinical questionnaire, PTSD data set

```{r}
rm(list = ls())
```


**LOAD PACKAGE** 
```{r}
# psych package
# install.packages('psych')
library(psych)

# MVN package for multivariate normality test
# install.packages("MVN")
library(MVN)

# nFactors package 
# install.packages('nFactors')
library(nFactors)

#install.packages('GPArotation')
library(GPArotation)

library(ggplot2)

#install.packages("GGally")
library(GGally)

# install.packages("PerformanceAnalytics")
library(PerformanceAnalytics)

# install.packages('xlsx')
# library(xlsx)
```

**LOAD DATA**
```{r}
tb = read.csv("D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/question scores EFA_09152018.csv", header = TRUE)

# exclude those who have incomplete data 
exclude = c(117, 1285, 100, 102, 50, 3) 

# find out who are remitted PTSD
remitted = tb$id[tb$group=='R']

```

**SELECT SUBJECT**  
Choice 1. include only control and PTSD, but also include women (should be 59 participants)
```{r}
tb2faGain <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]

tb2faLoss <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]
```

Choice 2. include only control and PTSD, (should be 55 participants)
```{r}
tb2faGain <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                & tb$isMale==1 & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]

tb2faLoss <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                & tb$isMale==1 & !is.element(tb$id,exclude) 
                & !is.element(tb$id,remitted),]
```

Choice 3. include control, PTSD, remitted PTSD (should be 69 participants)
```{r}
tb2faGain <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
            & tb$isMale==1 & !is.element(tb$id,exclude),]

tb2faLoss <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                & tb$isMale==1 & !is.element(tb$id,exclude),]
```

Display column names of data table, make sure to check if the right columns are selected to do PCA
```{r}
names(tb2faGain)
```

**Prepare data for PCA**  

Organize into a seperate array, reorder items so that the correlation plots look nicer
```{r}
array2fa <- tb2faGain[,c(35:39,34,47:48,54,49,46)] # 11 questionnaitre
names(array2fa)
colnames(array2fa) = c("CAPS-ReExp","CAPS-Avoid","CAPS-Numb","CAPS-DysA","CAPS-AnxA","BDI", 
                       "STAI-1","STAI-2","DES","CES","CTQ")
```

Organize into a seperate array
```{r}
array2fa <- tb2faGain[,c(34:39, 46, 47:49, 54)] # 11 questionnaitre
colnames(array2fa) = c("BDI", "CAPS-ReExp","CAPS-Avoid","CAPS-Numb","CAPS-DysA","CAPS-AnxA",
                       "CTQ","STAI-1","STAI-2","CES","DES")
```

Test multivariate normality
```{r}
mvn(array2fa)
```

Visualize data structure
```{r}
cor.plot(array2fa, numbers = TRUE, gr = colorRampPalette(c("#2171B5","white","#B52127")), zlim = c(-1,1), main="11 Quest Scores", xlas=2)

# with adjust for multiple tests
cor.plot(array2fa, numbers = TRUE, adjust ="holm", main="11 Quest Scores", xlas=2)


chart.Correlation(array2fa,
                  histogram = TRUE,
                  method = 'pearson')

ggcorr(array2fa, low = "#2171B5", mid = "white", high = "#B52127", label = TRUE, label_size = 3, label_round = 2, label_alpha = TRUE, size = 3, hjust = 0.75, color = "grey50")


# pairs.panels(array2fa,pch='.')
```

Z-Score
```{r}
# generate z-scores for variable A using the scale() function
zscored <- scale(array2fa, center = TRUE, scale = TRUE)

```

**Do PCA**  
There are two functions to use, princomp and prcomp
```{r}
# pcaresult <- princomp(zscored)
pcaresult <- prcomp(array2fa, retx = TRUE, center = TRUE, scale. = TRUE)
summary(pcaresult)
```
Plot variance explained
```{r}
# calculate variace from eigenvalues
eigs <- pcaresult$sdev^2
var_explain <- eigs/sum(eigs)
plot(var_explain, type = "b")

comp_id = c(1:11)
var_plot <- data.frame(comp_id, var_explain)

# plot variance explained
ggplot(var_plot, aes(x = comp_id, y = var_explain)) +
  geom_point() + geom_line() +
  scale_x_continuous(breaks = c(1:11)) +
  theme_classic()

# plot cumulative variance
for (i in c(1:nrow(var_plot))) {
  var_plot$var_cumul[i] <- sum(var_plot$var_explain[1:i])
}


# plot cumulative variance explained
ggplot(var_plot, aes(x = comp_id, y = var_cumul)) +
  geom_point(size = 3) + geom_line() +
  scale_x_continuous(breaks = c(1:11)) +
  theme_classic() +
  theme(text = element_text(size=12), axis.line = element_line(size = 1),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text = element_text(colour = "black", size = 10)) +
  labs(x="Number of components", y="Cumulative variance explained")

# another way to plot eigenvalues
plot(pcaresult)
```

Plot loadings
```{r}
# plot loadings
# pcaload <- pcaresult$loadings[,]
pcaload <- pcaresult$rotation[,]
# barplot(pcaload[,1], main="Component 1", ylab="Loadings", las=2)
# barplot(pcaload[,2], main="Component 2", ylab="Loadings", las=2)
# barplot(pcaload[,3], main="Component 3", ylab="Loadings", las=2)

load_plot <- data.frame(comp_id, pcaload[,1], pcaload[,2], pcaload[,3])
colnames(load_plot) <- c('comp_id', 'comp1', 'comp2', 'comp3')

# bar plot of loadings
ggplot(load_plot, aes(x = comp_id, y = comp1)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = c(1:11), labels = rownames(load_plot)) +
  scale_y_continuous(limits = c(0,0.4), breaks = seq(0,0.4,0.05)) +
  theme_classic() +
  theme(text = element_text(size=12), axis.line = element_line(size = 1),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text.x = element_text(angle = 90),
        axis.text = element_text(colour = "black", size = 10)) +
  labs(y="Loadings", x="") +
  ggtitle('Component1')

ggplot(load_plot, aes(x = comp_id, y = comp2)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = c(1:11), labels = rownames(load_plot)) +
  scale_y_continuous(limits = c(-0.4,0.6), breaks = seq(-0.4,0.6,0.2)) +
  theme_classic() +
  theme(text = element_text(size=12), axis.line = element_line(size = 1),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text.x = element_text(angle = 90),
        axis.text = element_text(colour = "black", size = 10)) +
  labs(y="Loadings", x="") +
  ggtitle('Component2')

ggplot(load_plot, aes(x = comp_id, y = comp3)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = c(1:11), labels = rownames(load_plot)) +
  # scale_y_continuous(limits = c(-0.4,1), breaks = seq(-0.4,0.85,0.2)) +
  theme_classic() +
  theme(text = element_text(size=12), axis.line = element_line(size = 1),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text.x = element_text(angle = 90),
        axis.text = element_text(colour = "black", size = 10)) +
  labs(y="Loadings", x="") +
  ggtitle('Component3')

```

Get rotated values (component scores), and plot on PC dimensions
```{r}
# rotated values
score = pcaresult$x

View(score)
typeof(score)
# change it to data frame
score_tb <- as.data.frame.array(score)

View(score_tb)
View(tb2faGain)

score_tb$id <- tb2faGain$id
score_tb$group <- tb2faGain$group


# plot PC distribution
ggplot(score_tb, aes(x = PC1, color = group, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.6, size =1, bins = 25) +
  scale_color_manual(values = c("olivedrab","darkorange3","grey50")) +
  scale_fill_manual(values = c("olivedrab","darkorange3","grey50")) +
  theme_classic() +
  theme(text = element_text(size=12), axis.line = element_line(size = 1),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text = element_text(colour = "black", size = 10)) +
  labs(x="PC1", y="Count of participant")

ggplot(score_tb, aes(x = PC2, color = group, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.6, size =1, bins = 25) +
  scale_color_manual(values = c("olivedrab","darkorange3","grey50")) +
  scale_fill_manual(values = c("olivedrab","darkorange3","grey50")) +
  theme_classic() +
  theme(text = element_text(size=12), axis.line = element_line(size = 1),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text = element_text(colour = "black", size = 10)) +
  labs(x="PC2", y="Count of participant")

ggplot(score_tb, aes(x = PC3, color = group, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.6, size =1, bins = 25) +
  scale_color_manual(values = c("olivedrab","darkorange3","grey50")) +
  scale_fill_manual(values = c("olivedrab","darkorange3","grey50")) +
  theme_classic() +
  theme(text = element_text(size=12), axis.line = element_line(size = 1),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text = element_text(colour = "black", size = 10)) +
  labs(x="PC3", y="Count of participant")

# test if Control and PTSD differ in PC3 (trauma exposure)
t.test(score_tb$PC2[score_tb$group=='C'], score_tb$PC2[score_tb$group=='P'])


# plot participants in PC space
ggplot(score_tb, aes(x=PC1, y=PC2, color=group)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("olivedrab","darkorange3","grey50")) +
  scale_fill_manual(values = c("olivedrab","darkorange3","grey50")) +
  # theme_classic() +
  theme(text = element_text(size=12),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text = element_text(colour = "black", size = 10))

ggplot(score_tb, aes(x=PC2, y=PC3, color=group)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("olivedrab","darkorange3","grey50")) +
  scale_fill_manual(values = c("olivedrab","darkorange3","grey50")) +
  # theme_classic() +
  theme(text = element_text(size=12),
        axis.ticks = element_line(colour = "black", size = 1),
        axis.text = element_text(colour = "black", size = 10))


```

**Save PCA output**  
Combine component scores into data
```{r}
# combine component into data
tb2faGain$comp1 = score[,1]
tb2faGain$comp2 = score[,2]
tb2faGain$comp3 = score[,3]

tb2faLoss$comp1 = score[,1]
tb2faLoss$comp2 = score[,2]
tb2faLoss$comp3 = score[,3]
```

Save data
```{r}
# save data
write.csv(tb2faGain,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score gain_noRemitted_08072019.csv")
write.csv(tb2faLoss,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score loss_noRemitted_08072019.csv")
```

**Caculate the components for remitted PTSD subjects**  
If excluding females
```{r}
# select remitted subjects only
tb2faGainRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                        & tb$isMale==1 & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]

tb2faLossRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                        & tb$isMale==1 & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]
```

If including females
```{r}
# select remitted subjects only
tb2faGainRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 1
                        & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]

tb2faLossRemitted <- tb[tb$isExcluded_behavior == 0 & tb$isGain == 0
                        & !is.element(tb$id,exclude) 
                        & is.element(tb$id,remitted),]

```

Calculate the components
```{r}
scoreRemitted = scale(tb2faGainRemitted[,c(35:39,34,47:48,54,49,46)], center= pcaresult$center) %*% pcaload
```

Combine component score into data
```{r}
tb2faGainRemitted$comp1 = scoreRemitted[,1]
tb2faGainRemitted$comp2 = scoreRemitted[,2]
tb2faGainRemitted$comp3 = scoreRemitted[,3]

tb2faLossRemitted$comp1 = scoreRemitted[,1]
tb2faLossRemitted$comp2 = scoreRemitted[,2]
tb2faLossRemitted$comp3 = scoreRemitted[,3]

```

Save data
```{r}

write.csv(tb2faGainRemitted,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score gain_Remitted_08072019.csv")
write.csv(tb2faLossRemitted,"D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral/pca score loss_Remitted_08072019.csv")

```





When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
