---
title: "subjective_value"
author: ruonan jia
date: 06.14.2019
---

---
load packages
```{r}
library(ggplot2)
library(ez)
library(psych)
library(lme4)
```

---
functions
```{r}
data_summary <- function(data, varname, groupnames){
  # Function to calculate the mean and the standard error
  # for each group
  #+++++++++++++++++++++++++
  # data : a data frame
  # varname : the name of a column containing the variable
  #to be summariezed
  # groupnames : vector of column names to be used as
  # grouping variables
  
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE)/sqrt(length(x[[col]])))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}

```

---
load data
```{r}
behav_path <- "D:/Ruonan/Projects in the lab/VA_RA_PTB/Clinical and behavioral"
setwd(behav_path)

sv <- read.csv('sv_trial_day1day2.csv', header = TRUE)
load("data_all_noFemale_day1day2_05122019.rda")
pcl <- read.csv('pcl.csv',header = TRUE)
demo <- read.csv('ed_income.csv', header = TRUE)
demo$education <- factor(demo$education, order = TRUE, levels = c(1,2,3,4,5,6))
demo$income <- factor(demo$income, order = TRUE, levels = c(1,2,3,4,5,6,7,8,9,10))

# combine data
tball_pcl <-  merge(tball, pcl, by = 'id')
tb_all <- merge(tball_pcl, demo, by = 'id')
tb <- tb_all[tb_all$isExcluded_imaging==0 & tball$isMale==1,]

```

---
plot SV distribution of a single subject
```{r}
id2plot <- 3
isDay12plot <- 1
isGain2plot <- 0
isRisk2plot <- 1

if (isRisk2plot == 1){
  sv2plot <- sv[sv$id == id2plot & sv$isDay1 == isDay12plot & sv$isGain == isGain2plot & sv$ambig == 0,]
} else {
  sv2plot <- sv[sv$id == id2plot & sv$isDay1 == isDay12plot & sv$isGain == isGain2plot & sv$ambig > 0,] 
}

ggplot(sv2plot, aes(x=sv)) +
  geom_histogram(bins = 30) + 
  ggtitle(paste('Subject', id2plot, ', SV Histogram. isGain=', isGain2plot, ', isDay1=', isDay12plot, ', isRisk=', isRisk2plot))

```

---
calculate the mean and variance of SV in each decision domain for each subject
```{r}
# get subject id
subject_id <- unique(tb$id)
subject_id

# SV summary (min, max, variance), day1/day2, gain/loss, risk/ambiguity
row_idx <- 0
sv_sum <- data.frame(id=integer(),
                 isDay1=integer(),
                 isGain=integer(),
                 isRisk=integer(),
                 sv_max=double(),
                 sv_min=double(),
                 sv_mean=double(),
                 sv_sd=double(),
                 sv_ref=double())

for (id2sum in subject_id) {
  for (isDay12sum in c(1,0)) {
    for (isGain2sum in c(1,0)) {
      for (isRisk2sum in c(1,0)) {
        if (isRisk2sum == 1) {
          sv2sum <- sv[sv$id == id2sum & sv$isDay1 == isDay12sum & sv$isGain == isGain2sum & sv$ambig == 0,] 
        } else {
          sv2sum <- sv[sv$id == id2sum & sv$isDay1 == isDay12sum & sv$isGain == isGain2sum & sv$ambig > 0,]
        }
        
        row_idx <- row_idx + 1
        
        sv_sum[row_idx,]$id <- id2sum
        sv_sum[row_idx,]$isDay1 <- isDay12sum
        sv_sum[row_idx,]$isGain <- isGain2sum
        sv_sum[row_idx,]$isRisk <- isRisk2sum
        
        sv_sum[row_idx,]$sv_max <- max(sv2sum$sv)
        sv_sum[row_idx,]$sv_min <- min(sv2sum$sv)
        sv_sum[row_idx,]$sv_mean <- mean(sv2sum$sv)
        sv_sum[row_idx,]$sv_sd <- sd(sv2sum$sv)
        sv_sum[row_idx,]$sv_ref <- mean(sv2sum$sv_ref)
      }
    }
  }
}
```


---
merge data frame
```{r}
sv_sum$id <- as.factor(sv_sum$id)

grouping <- data.frame(id = tb[tb$isDay1 == 1 & tb$isGain == 1,]$id,
                      isExcluded_behavior = tb[tb$isDay1 == 1 & tb$isGain == 1,]$isExcluded_behavior,
                      isExcluded_imaging = tb[tb$isDay1 == 1 & tb$isGain == 1,]$isExcluded_imaging,
                      isMale = tb[tb$isDay1 == 1 & tb$isGain == 1,]$isMale,
                      age = tb[tb$isDay1 == 1 & tb$isGain == 1,]$age,
                      kbit = tb[tb$isDay1 == 1 & tb$isGain == 1,]$kbit,
                      group = tb[tb$isDay1 == 1 & tb$isGain == 1,]$group)

sv_sum_group <- merge(sv_sum, grouping, by = 'id')

```

---
compare sv range between PTSD and control
```{r}
isDay12plot <- 0
isGain2plot <- 0
isRisk2plot <- 0
group2plot <- c('P', 'C')

sv_sum2plot <- sv_sum_group[sv_sum_group$isExcluded_imaging == 0 & match(sv_sum_group$group, group2plot) & sv_sum_group$isDay1 == isDay12plot & sv_sum_group$isGain == isGain2plot & sv_sum_group$isRisk == isRisk2plot,]
 

ggplot(sv_sum2plot[sv_sum2plot$sv_mean > -30,], aes(x=sv_mean, color=group, fill = group)) +
  geom_histogram(bins = 30, position = 'identity', alpha = 0.5) + 
  ggtitle(paste('SV Histogram. isGain=', isGain2plot, ', isDay1=', isDay12plot, ', isRisk=', isRisk2plot)) +
  theme(plot.title = element_text(size = 12))


sv_sum2plot[sv_sum2plot$sv_mean>-30 & sv_sum2plot$sv_mean< -10, ]$id
```




Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
